Boosting Higgs Pair Production in the $b\bar{b}b\bar{b}$ Final State with Multivariate Techniques

### 2015-12-08 paper/sec-mva.tex
more work on the MVA section

The main limitation of our analysis so far is the fact that
pile-up (PU) effects have not been accounted for
not accounted for.
%
However, any realistic feasibility study at the HL-LHC, specially
one 
such the present that relies on the careful modeling
of kinematic distributions,
requires assessing the robustness of the conclusions
with respect to the inclusion of PU.
%
In this section, we show how our qualitative results, in particular
the enhanced signal significance provided by the MVA, are robust
in the presence of  PU.

First of all, we describe the simulation of the 
PU conditions  expected at the HL-LHC, and discuss the settings of
the PU subtraction strategy that we adopt.
%
Then we present the validation of the PU subtraction,
and compare a number of distributions, including substructure variables,
with and without PU.
%
Finally we revisit the MVA analysis of Sect.~\ref{sec:mva}, and
show that our qualitative conclusions are robust
in the presence of realistic PU conditions, finding
a combined
signal significance of $S/\sqrt{B}\simeq 4$.

The same settings of the cut-based analysis,
Sects.~\ref{sec:analysis} and~\ref{sec:results}, and
the subsequent MVA optimization, Sect.~\ref{sec:mva}, have
been applied to the samples with PU, $\la n_{\rm PU}\ra=80$,
and SK subtraction.
%
The corresponding version of Table~\ref{tab:cutflow_noPU_1}, the 
cut-flow for the cross-sections and signal significance at the various
steps of the cut-based analysis, 
in the case
of PU subtracted with SK, for $\la n_{\rm PU}\ra=80$,
is presented in Table~\ref{tab:cutflow_SKPU80_1}.
%
One immediate observation is that
at the end of the cut-based analysis,
the signal significance have been  degraded
as compared to the case without PU.
%
We now obtain values for $S\sqrt{B}$ of 0.4, 0.3 and 0.9, in the resolved,
intermediate and boosted categories, respectively, to be compared
with the corresponding values without PU, namely 1.6, 1.6 and 2.7.
%
Therefore, the pre-MVA signal significance is degraded by a factor 4 in
the resolved and the intermediate categories,
and by a factor 3 in the boosted category.
%
The fact that the latter is the  least affected one
can be understood from the previous comparison
plots, which show that effects of PU contamination
are milder due to the  selection requirements
of the boosted category.

@@

In Table~\ref{table:cutflowMVA_PU}
we provide the  the number of signal and
    background events expected
    at the HL-LHC, as well as the
    corresponding values for $S/\sqrt{B}$ and $S/B$,
    for both the cut-based analysis (corresponding
    to $y_{\rm cut}=0$) and after the
    optimal MVA cut, for the simulations
    with $\la n_{\rm PU}\ra=80$
    and SK subtraction.
    %
    The same results without PU are those of
    Table.~\ref{table:cutflowMVA}.
    %
    From Table~\ref{table:cutflowMVA_PU} we see that, thanks
to the MVA, we can improve the signal significance and the
signal over background ratio from 0.35 and $4\cdot 10^{-4}$
(0.25 and $1\cdot 10^{-4}$) up to 2.9 and 0.034 (2.5 and 0.15)
in the boosted (resolved) category.
%
The intermediate category exhibits a rather lower value of $S/\sqrt{B}$
for the optimal $y_{\rm cut}$ value, around 1.1.
%
We thus conclude that, as was already the case
without PU,
the boosted category is the most promising
one for the study of Higgs pair production in the $b\bar{b}b\bar{b}$
final state
at the HL-LHC, also when PU is included.
%
However, important information will also be obtained from
the resolved category, which has a signal significance
only slightly smaller than the boosted case.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}[t]
  \centering
  \begin{tabular}{|c|l|c|c|c|c|}
     \multicolumn{6}{|c|}{$\la n_{\rm PU}\ra=80$+SK} \\
    \hline
    Category  &   &  $N_{\rm ev}$ signal &  $N_{\rm ev}$ back  &  $S/\sqrt{B}$ & $S/B$ \\ 
    \hline
    \hline
    \multirow{2}{*}{Boosted} &  $y_{\rm cut}=0$  & 360   &  $1.4\cdot 10^6$ & 0.31   &
     $3\cdot 10^{-4}$  \\
    &  $y_{\rm cut}=0.82$ &  250 & 7300  & 2.9    & 0.034  \\
    \hline
    \hline
    \multirow{2}{*}{Intermediate} &  $y_{\rm cut}=0$  &  150  & $9\cdot 10^5$    & 0.15    &
     $2\cdot 10^{-4}$ \\
    &  $y_{\rm cut}=0.80$ & 50 & 2600  &  1.1   & 0.02 \\
    \hline
    \hline
    \multirow{2}{*}{Resolved} &  $y_{\rm cut}=0$  &  1100  & $8.5\cdot 10^6$
    & 0.25    &  $1.3\cdot 10^{-4}$  \\
    &  $y_{\rm cut}=0.65$ & 430  & $3\cdot 10^4$  &  2.5   & 0.15  \\
    \hline
      \end{tabular}
  \caption{\small Same as Table~\ref{table:cutflowMVA}, now for the case
    of PU+SK, with $\la n_{\rm PU}\ra=80$.
        \label{table:cutflowMVA_PU}
  }
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Another important result from Table~\ref{table:cutflowMVA_PU} is that,
for the optimal MVA cut, the signal over background ratio
is not unfeasibly small: we obtain that $S/B$ is 3.4\% (1.5\%)
in the boosted (resolved) categories.
%
This is indicates that while this measurement is still very challenging,
requiring a careful extraction from the data of the QCD
background, it is certainty within reach.
%
Our analysis also indicates a possible way forward to
further increase $S/B$: reducing the background from light and charm
fakes, see Table~\ref{tab:cutflow_SKPU80_1}.

Let us now discuss in more detail the results of the MVA
classification, by showing various estimators as function
of the cut $y_{\rm cut}$ in the ANN discriminant.
%
In Fig.~\ref{fig:nev2_PU}
we show the number of signal and background events that
are expected at the HL-LHC as a function of
$y_{\rm cut}$ -
the corresponding results without PU are those of
Fig.~\ref{fig:nev2}.
%
As we can see, the general trends are
similar as those without PU.
%
Comparing Figs.~\ref{fig:nev2} and~\ref{fig:nev2_PU}, we also observe
that the intermediate category is substantially degraded once PU effects
are accounted for, the reason being  sizable
a reduction in the number of signal
events that are now assigned to this category.
%
This shows that the selection criteria
of the intermediate category are less
resilient against PU contamination,
as opposed to the resolved and boosted selections.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
\begin{center}
\includegraphics[width=0.65\textwidth]{plots/nev2_SKPU80.pdf}
\caption{\small Same as Fig.~\ref{fig:nev2} in the
case of events with PU, for
 $\la n_{\rm PU}\ra=80$ 
  and SK subtraction.
}
\label{fig:nev2_PU}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In Fig.~\ref{fig:sb_mva_PU} we show the signal significance,
$S/\sqrt{B}$, as well as the signal over background ratio,
$S/B$, including now the effects of PU.
%
The corresponding results in the case without PU were shown in
Fig.~\ref{fig:sb_mva}.
%
As can be seen, the MVA-driven enhancement is robust in the
presence of PU.
%
While unsurprisingly there us a degradation as compared to the case wo PU,
we still manage to achieve a signal significance of
around $S/\sqrt{B}\simeq 2.5$ separately for the boosted and resolved
categories.
%
On the other hand, the intermediate category becomes irrelevant.
%
We conclude that the qualitative conclusions drawn in
Sect.~\ref{sec:mva} are robust in the presence
of realistic PU effects.
%
Since no specific effort has been performed to
optimize PU subtraction, for example by tuning the value
of the patch length $a$ in {\tt SoftKiller}, we believe that
there is
still room for improvement in this respect.

@@

It is important to quantify which of the input variables
to the MVA carry the highest discrimination power
in the case of PU,
and compare it with the corresponding
results without PU.
%
We use the same estimator as in Sect.~\ref{sec:signalsignificance},
namely the sum
of the absolute value of all the weights connected to a given
input neuron $i$, Eq.~(\ref{eq:totweight}).
%
The results for the resolved and boosted categories are shown
on Fig.~\ref{fig:nnweights_PU}.
%
As we can see by comparing to the corresponding
results without PU in Fig.~\ref{fig:nnweights}, 
in the boosted category, the discrimination power of the invariant
mass of Higgs candidates is decreased and that of the various substructure
variables, in particular $C_2^{(\beta)}$ and
$D^{(\beta)}$, is conversely
increased.
%
This reflects the fact that these substructure variables are
relatively robust against PU contamination.
%
In addition, also the $p_T$ distributions of the AK03
subjets within the large-$R$
jets carry important information.
%
In the case of the resolved category,  the highest
values of the ANN weights without PU
were found for the $p_T$ of the leading
Higgs and for the Higgs invariant masses.
%
This is also true in the case with PU, but now the rapidity difference
between the two Higgs candidates, $\Delta y_{hh}$ increases its
importance, again reflecting that this variable is relatively
insensitive to PU effects.
%

@@

At this point, we can finally combine the three categories, resolved,
intermediate and boosted, and determine the overall signal
significance taking into account all event topologies.
%
Adding in quadrature the signal significance from the three
(exclusive) categories for the corresponding
optimal value of $y_{\rm cut}$, our final combined result, accounting
for PU effects, is
\be
\lp \frac{S}{\sqrt{B}}\rp_{\rm tot} \simeq 4.0 \, ,
\ee
to be compared to the value $\lp S/\sqrt{B}\rp_{\rm tot} \simeq 8.5$
that was obtained in the case without PU.
%
We conclude that, even when realistic PU effects are accounted
for, at the HL-LHC the measurement of
Higgs pair production in the $b\bar{b}b\bar{b}$ final state should be 
well the threshold for evidence, and even meeting the
requirement for discovery might be within reach.
%

In summary, in this section we have demonstrated the robustness
of our results under the presence of PU conditions as those
expected at the HL-LHC.
%
With a signal significance of $S/\sqrt{B}\sim 4$, is clear that
the $b\bar{b}b\bar{b}$ final state can be used to claim evidence
for Higgs pair production at the HL-LHC without the need
to combine with any other final state.
%
Our analysis certainly contains
room for improvement, in particular by means of a tailored
study dedicated to optimize PU subtraction in this process.
%
It will be important now to quantify the precision of
a extraction of the Higgs trilinear coupling $\lambda$ using
the techniques presented here, but this is left for future work.

>>>>

In this section we study how the MVA results are modified when
signal and background events are embedded in a high-PU environment.
%
The simulation of PU and the corresponding subtraction with
{\tt SoftKiller} have been performed as described
in Sect.~\ref{sec:pileup}.
%
Following this, the cut-based analysis and the subsequent
MVA optimization have been applied using the same
settings as in the case without PU.
%
In Table~\ref{tab:cutflow_SKPU80_1} we provide the results
of the cut-based analysis cut-flow, for the case with $\la n_{\rm PU}\ra=80$+SK.
%
It is clear that the pre-MVA 
signal significance have been seriously degraded
as compared to the case without PU.
%
We now find values for $S/\sqrt{B}$ of 0.4, 0.3 and 0.9, in the resolved,
intermediate and boosted categories, respectively, to be compared
with the corresponding values without PU, namely 1.6, 1.6 and 2.7.
%

@@

Following the cut-based analysis, the surviving signal and background
MC events in the case of PU are used to re-train the MVA.
%
In Table~\ref{table:cutflowMVA_PU}
we provide  the number of signal and
    background events expected
    at the HL-LHC with $\mathcal{L}=3$ ab$^{-1}$,
    as well as the
    corresponding values for $S/\sqrt{B}$ and $S/B$,
    for both the cut-based analysis (corresponding
    to $y_{\rm cut}=0$) and after the
    optimal MVA cut, for the simulations
    with $\la n_{\rm PU}\ra=80$
    and SK subtraction.
    %
    From Table~\ref{table:cutflowMVA_PU} we observe that, thanks
to the MVA, the signal significance and the
signal over background ratio are improved from 0.35 and $4\cdot 10^{-4}$
(0.25 and $1\cdot 10^{-4}$) up to 2.9 and 0.034 (2.5 and 0.15)
in the boosted (resolved) category.
%
The intermediate category, on the other hand, exhibits a rather lower post-MVA value of $S/\sqrt{B}$, around
1.1.
%
We thus conclude that, as was already the case
without PU,
the boosted category is the most promising
one for the study of Higgs pair production in the $b\bar{b}b\bar{b}$
final state
at the HL-LHC, also for the case of a high-PU environment, and that
important information will also be obtained from
the resolved category, which has a comparable significance.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}[t]
  \centering
  \begin{tabular}{|c|l|c|c|c|c|}
        \hline
     \multicolumn{6}{|c|}{$\la n_{\rm PU}\ra=80$+SK} \\
     \hline
         \hline
    Category  &   &  $N_{\rm ev}$ signal &  $N_{\rm ev}$ back  &  $S/\sqrt{B}$ & $S/B$ \\ 
    \hline
    \hline
    \multirow{2}{*}{Boosted} &  $y_{\rm cut}=0$  & 360   &  $1.4\cdot 10^6$ & 0.31   &
     $3\cdot 10^{-4}$  \\
    &  $y_{\rm cut}=0.82$ &  250 & 7300  & 2.9    & 0.034  \\
    \hline
    \hline
    \multirow{2}{*}{Intermediate} &  $y_{\rm cut}=0$  &  150  & $9\cdot 10^5$    & 0.15    &
     $2\cdot 10^{-4}$ \\
    &  $y_{\rm cut}=0.80$ & 50 & 2600  &  1.1   & 0.02 \\
    \hline
    \hline
    \multirow{2}{*}{Resolved} &  $y_{\rm cut}=0$  &  1100  & $8.5\cdot 10^6$
    & 0.25    &  $1.3\cdot 10^{-4}$  \\
    &  $y_{\rm cut}=0.65$ & 430  & $3\cdot 10^4$  &  2.5   & 0.015  \\
    \hline
      \end{tabular}
  \caption{\small Same as Table~\ref{table:cutflowMVA}, now for the case
    of PU+SK, with $\la n_{\rm PU}\ra=80$.
        \label{table:cutflowMVA_PU}
  }
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Another important result from Table~\ref{table:cutflowMVA_PU} is that,
for the optimal MVA cut, the signal over background ratio
is not unfeasibly small: we obtain that $S/B$ is 3.4\% (1.5\%)
in the boosted (resolved) categories.
%
This is indicates that while this measurement is still very challenging,
requiring a careful extraction from the data of the QCD
background, it is certainty within reach.
%
As in the case without PU, 
this analysis  indicates that a substantial increase in both
$S/B$ and $S/\sqrt{B}$ could be achieved by reducing the background component
that arises from the mis-identification of light and charm
fakes, compare for in particular the columns $S/\sqrt{B_{\rm tot}}$
and  $S/\sqrt{B_{\rm 4b}}$ in  Table~\ref{tab:cutflow_SKPU80_1}.

In Fig.~\ref{fig:nev2_PU}
we show the number of signal and background events that
are expected at the HL-LHC as a function of
$y_{\rm cut}$ and the corresponding ROC curve
%
Comparing Figs.~\ref{fig:nev2} and~\ref{fig:nev2_PU}, we observe
that the intermediate category is substantially degraded once PU effects
are accounted for, the reason being  sizable
a reduction in the number of signal
events that are now assigned to this category.
%
This shows that the selection criteria
of the intermediate category are less
resilient against PU contamination,
as opposed to the resolved and boosted selections.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
  \begin{center}
    \includegraphics[width=0.49\textwidth]{plots/roc_SKPU80.pdf}
\includegraphics[width=0.49\textwidth]{plots/nev2_SKPU80.pdf}
\caption{\small Same as Fig.~\ref{fig:nev2} in the
case of events with PU, for
 $\la n_{\rm PU}\ra=80$ 
  and SK subtraction.
}
\label{fig:nev2_PU}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In Fig.~\ref{fig:sb_mva_PU} we show the signal significance,
$S/\sqrt{B}$, as well as the signal over background ratio,
$S/B$, including now the effects of PU.
%
The corresponding results in the case without PU were shown in
Fig.~\ref{fig:sb_mva}.
%
As can be seen, the MVA-driven enhancement is robust in the
presence of PU.
%
While unsurprisingly there us a degradation as compared to the case wo PU,
we still manage to achieve a signal significance of
around $S/\sqrt{B}$ between 2.5 and 3.0
separately for the boosted and resolved
categories.
%
On the other hand, the intermediate category becomes marginal.
%
We conclude that the qualitative conclusions drawn in
Sect.~\ref{sec:mva} are robust in the presence
of realistic PU effects.
%
Since no specific effort has been performed to
optimize PU subtraction, for example by tuning the value
of the patch length $a$ in {\tt SoftKiller}, we believe that
there is
still room for improvement in this respect.

@@

It is important to quantify which of the input variables
to the MVA carry the highest discrimination power
in the case of PU,
and compare it with the corresponding
results without PU.
%
We use the same estimator as in Sect.~\ref{sec:signalsignificance},
namely the sum
of the absolute value of all the weights connected to a given
input neuron $i$, Eq.~(\ref{eq:totweight}).
%
The results for the resolved and boosted categories are shown
on Fig.~\ref{fig:nnweights_PU}.
%
As we can see by comparing to the corresponding
results without PU in Fig.~\ref{fig:nnweights}, 
in the boosted category, the discrimination power of the invariant
mass of Higgs candidates is decreased and that of the various substructure
variables, in particular $C_2^{(\beta)}$ and
$D^{(\beta)}$, is conversely
increased.
%
This reflects the fact that these substructure variables are
relatively robust against PU contamination.
%
In addition, also the $p_T$ distributions of the AKT03
subjets within the large-$R$
jets carry important information.
%
In the case of the resolved category,  the highest
values of the ANN weights without PU
were found for the $p_T$ of the leading
Higgs and for the Higgs invariant masses.
%
This is also true in the case with PU, but now the rapidity difference
between the two Higgs candidates, $\Delta y_{hh}$ increases its
importance, again reflecting that this variable is relatively
insensitive to PU effects.
%

@@

We can finally combine the three categories, resolved,
intermediate and boosted, and determine the overall signal
significance, finding
\be
\lp \frac{S}{\sqrt{B}}\rp_{\rm tot} \simeq 4.0 \, ,
\ee
to be compared to the value $\lp S/\sqrt{B}\rp_{\rm tot} \simeq 8.5$
that was obtained in the case without PU.
%
We conclude that, even when realistic PU effects are accounted
for, at the HL-LHC the measurement of
Higgs pair production in the $b\bar{b}b\bar{b}$ final state should be 
well the threshold for evidence, and even meeting the
requirements for discovery might be within reach provided it is possible
to suppress the background from light and charm jet mis-identifications.

For the Run II luminosity, $\mathcal{L}=300$ fb$^{-1}$, one finds
instead $\lp S/\sqrt{B}\rp_{\rm tot} \simeq 1.3$, which increases up to
$\lp S/\sqrt{B}\rp_{\rm tot} \simeq 2.6$ if only the QCD $4b$ background is included.
%
This indicates that, interestingly, accessing Higgs pair production at Run II,
while still extremely challenging, does not seem to be completely impossible.
%
In addition, at Run II the average number of PU collisions is substantially
reduced as compared to the HL-LHC environment.
%
Therefore, it will be interesting to see if improvements in experimental $b$-jet
reconstruction and fake rejection at Run II are able to fulfill this
potential.
